A Constraint Solving Approach to Parikh Images of Regular Languages

### 2024-02-29 evaluation.tex
Poke the introduction
Closes: #86
Closes: #88
Closes: #82
Closes: #80

We obtain the results by executing \Ostrich{}~1.3 with CA-Str using
the same benchmarking infrastructure (the StarExec cluster)
and configuration that ran SMT-COMP~2023, combining our new results for CA-Str with the
published results of SMT-COMP~2023.
\footnote{StarExec Job ID 59410, \enquote{SMT-COMP 2023 single query final}, 
available at \url{https://www.starexec.org/starexec/secure/details/job.jsp?id=59410}.} 
Where avaiable, we use the revised, out-of
competition version of the results for solvers with bugs discovered during
the competition, including \Ostrich{} and Z3-Noodler. 
\footnote{StarExec Job ID 59668, \enquote{SMT-COMP 2023 single query final fixed} available at
\url{https://www.starexec.org/starexec/secure/details/job.jsp?id=59668}.}
Note that Z3-Noodler abstained on 70~instances, and thus has a lower total number of results.
We used the parallel results for all solvers.

@@

These results are expected. It is known that automata-based string solvers (\Ostrich{}, Z3-Noodler) 
tend to perform better on unsatisfiable than on satisfiable benchmarks, compared to solvers
that do not utilize automata and directly work on regular expressions (cvc5, z3alpha).
This is not unexpected, as the computation of automata representations of regular 
constraints can be expensive, and might be unnecessary for satisfiable formulas.
In addition, the algorithm in \Ostrich{} has stronger theoretical completeness
guarantees as cvc5 and Z3-Str, and ensuring completeness often has an adverse
effect on performance in practice~\cite{ostrich,Z3-str,cvc5}.

The results show that \Calculus{} can be used to enhance the performance of an
automaton-based string solver. Moreover, the cactus plot in \cref{tab:solve-status-smt-comp} 
illustrates that
CA-Str is immediately useful, boosting the results for the \Ostrich{} portfolio
even at the first datapoint. These results should be considered preliminary,
however, as we believe that a deeper integration of \Catra{} into string solvers
can lead to significant performance gains. In particular, the integration layer 
is currently too shallow to allow \Ostrich{} to learn clauses generated by \Catra{}
and additionally incurs subprocess-spawning overhead and others, as well as overhead
from serialising and deserialising the current Parikh automata problem into \Catra{}'s
input format.

@@

The most obvious threat to validity would be an unsound implementation. To
address this we have validated all reported solutions made by \Calculus{} with
\Nuxmv{}. A previous version contained a race with random restarts during
product materialisation causing non-deterministic unsoundness in \numprint{0.7}
of instances. Since addressing that bug we have observed no further soundness
issues in \Catra{}. During extended evaluations by reviewer suggestion for the 
camera-ready version of this paper, however, one single erroneous \textsc{Unsat} 
response on the QF\_S track of 
SMT-COMP (\texttt{QF\_S\/20230329-automatark-lu\/instance08425.smt2}) was discovered.
The most likely cause is a misgenerated constraint in the interaction surface between \Ostrich{}
and \Catra{}, since we have robustly tested 
\Catra{} separately from \Ostrich{} and vice versa. Additionally, Z3-Noodler
also similarly has one incorrect response, even in the bugfixed version. For both of
these, we apply the same rule as SMT-COMP, counting the erroneous responses for Z3-Noodler
and \Ostrich{}+CA as error results.

The second threat to validity is our implementation of automata operations. As
\Calculus{} by design offloads some of the product computation work onto
\Princess{}, the baseline could be unfairly disadvantaged by a slow automata
product implementation. We believe this is not an issue since similar
performance issues with the baseline approach have been reported for string
solvers. Additionally, profiling shows that baseline spends most of its time in
\Princess{}, suggesting that the automaton implementation is not the bottleneck.
Finally, the difference in performance between \Calculus{} and baseline has been
robust under significant optimisation of the automata library, additionally
strengthening this thesis.

>>>>

We obtain the results by executing \Ostrich{}~1.3 with CA-Str using
the same benchmarking infrastructure (the StarExec cluster)
and configuration that ran SMT-COMP~2023, combining our new results for CA-Str with the
published results of SMT-COMP~2023.
\footnote{StarExec Job ID 59410, \enquote{SMT-COMP 2023 single query final}, 
available at \url{https://www.starexec.org/starexec/secure/details/job.jsp?id=59410}.} 
Where avaiable, we use the revised, out-of
competition version of the results for solvers with bugs discovered during
the competition, including \Ostrich{} and Z3-Noodler. 
\footnote{StarExec Job ID 59668, \enquote{SMT-COMP 2023 single query final fixed} available at
\url{https://www.starexec.org/starexec/secure/details/job.jsp?id=59668}.}
Note that Z3-Noodler abstained on \numprint{70}~instances, and thus has a lower total number of results.
We used the parallel results for all solvers.

@@

These results are expected. It is known that automata-based string solvers (\Ostrich{}, Z3-Noodler) 
tend to perform better on unsatisfiable than on satisfiable benchmarks, compared to solvers
that do not utilize automata and directly work on regular expressions (cvc5, z3alpha).
This is not unexpected, as the computation of automata representations of regular 
constraints can be expensive, and might be unnecessary for satisfiable formulas.
In addition, the algorithm in \Ostrich{} has stronger theoretical completeness
guarantees than cvc5 and Z3-Str, and ensuring completeness often has an adverse
effect on performance in practice~\cite{ostrich,Z3-str,cvc5}.

The results show that \Calculus{} can be used to enhance the performance of an
automaton-based string solver. Moreover, the cactus plot in \cref{tab:solve-status-smt-comp} 
illustrates that
CA-Str is immediately useful, boosting the results for the \Ostrich{} portfolio
even at the first datapoint (though marginally). These results should be considered preliminary,
however, as we believe that a deeper integration of \Catra{} into string solvers
can lead to significant performance gains. In particular, the integration layer 
is currently too shallow to allow \Ostrich{} to learn clauses generated by \Catra{}
and additionally incurs overhead from serialising and deserialising the current Parikh
automata problem into \Catra{}'s input format.

@@

The most obvious threat to validity would be an unsound implementation. To
address this we have validated all reported solutions made by \Calculus{} with
\Nuxmv{}. A previous version contained a race with random restarts during
product materialisation causing non-deterministic unsoundness in \numprint{0.7}
of instances. Additionally, both we and the artefact evaluation committee independently
discovered a soundness issue in CA-Str on one instance in QF\_S where the 
\texttt{QF\_S\slash{}20230329-automatark-lu\slash{}instance08425.smt2} instance was
incorrectly reported as \textsc{Unsat}. The underlying issue was that \Catra{} generated
a too broad blocking clause when encountering unsatisifiability stemming from 
an intermittent product accepting only the empty string during product materialisation. 
We implemented a fix for that bug and re-evaluated all benchmarks in SMT-COMP.

Since addressing those bugs we have observed no further soundness
issues in neither \Catra{} nor CA-Str. Benchmarking results for both runs are included
in the artefact of this paper, and show virtually identical performance characteristics.

The second threat to validity is our implementation of automata operations. As
\Calculus{} by design offloads some of the product computation work onto
\Princess{}, the baseline could be unfairly disadvantaged by a slow automata
product implementation. We believe this is not an issue since similar
performance issues with the baseline approach have been reported for other string
solvers, as well as for a previous implementation in \Ostrich{}. Additionally,
profiling shows that baseline spends most of its time in
\Princess{}, suggesting that the automaton implementation is not the bottleneck.
Finally, the difference in performance between \Calculus{} and baseline has been
robust under significant optimisation of the automata library, additionally
strengthening this thesis.

### 2024-02-29 introduction.tex
FO FO
Poke the introduction
Closes: #86
Closes: #88
Closes: #82
Closes: #80

It is possible to compute an existential Presburger formula describing the
Parikh image of any context-free language in linear
time~\cite{generate-parikh-image}; for the special case of regular languages,
this result was also stated in \cite{muscholl-linear}. While theoretically
elegant, this construction has several disadvantages, often making it
impractical for integration into algorithms. Firstly, the constructed Presburger
formula contains a linear number of existential quantifiers in the size of the
considered grammar, as well as complex Boolean structure, which is needed to
express the connectedness of sets of productions considered in the construction.
Eliminating those quantifiers to obtain a quantifier-free representation of the
Parikh image has exponential complexity~\cite{DBLP:conf/issac/Weispfenning97},
and is often impossible in reasonable time. Just solving the Parikh image
membership problem is NP-complete, as it corresponds to computing a satisfying
assignment of the existential Presburger formula, and taxing for solvers as
well~\cite{ostrich-plus}.

Secondly, in applications involving regular languages, it is typically necessary
to consider the Parikh image not only of a single automaton, but of the
intersection of multiple automata. This problem arises in string solvers in
particular, as conjunctions of string constraints lead to the computation of
length images of products (intersections) of regular languages represented as
finite automata. Applying the approach in~\cite{generate-parikh-image} would in
this case require the eager computation of the product before its length image,
and result in an existential Presburger formula of exponential size (in the
number of automata). In several instances we have observed while solving
real-world string constraints, the computation of the product of automata
exhausts the memory of any machine available due to the exponential blow-up in
size of the product, quickly becoming intractable as the number of automata in
the product increases. The current best published mitigation for this problem is
an over-approximation that works by approximating the Parikh image of a product
of automata to be the conjunction of the image of the individual automata of the
product \cite{approximate-parikh}. This approach only works for unsatisfiable
instances, and comes with a harsh penalty for satisfiable instances.

@@

The key ideas of \Calculus{} is a combination of problem-aware case splitting,
lazy enforcement of automata connectivity, and lazy computation of products. In
essence, we use constraint-based methods to efficiently enforce the constraints
associated with computing Parikh images of regular languages.

We implement \Calculus{} as a plug-in theory for the \Princess{} automated
theorem prover.
%
We additionally wrap the Parikh image solver as a stand-alone tool, \Catra.
\Catra{} also supports a variant of the approximate method of
\cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} distinct Parikh automata intersection problems
generated by \OstrichPlus{} when solving the PyEx string constraint benchmark
suite involving string length constraints \cite{pyex}.

>>>>

In applications involving regular languages, it is typically necessary
to consider the Parikh image not only of a single automaton, but of the
intersection of multiple automata. This problem arises in string solvers in
particular, as conjunctions of string constraints lead to the computation of
length images of products (intersections) of regular languages represented as
finite automata. The problem considered in this paper is thus:
{
    \centering
    \begin{tcolorbox}[colback=gray!5!white,colframe=gray!75!black,%
        title=Satisfiability of Parikh images of a product of automata,%
        width=0.8\linewidth]
        Given a finite set of deterministic, finite-state automata, $\Automaton_1, \ldots, \Automaton_n$
        over a finite, common, alphabet $\Set{a_1, \ldots, a_k}$,
        and a Presburger formula, $\phi(\VectorLiteral{x_1, \ldots, x_k})$, 
        decide if there is a word, $w$ accepted by the 
        product $\Automaton_1 \times \cdots \times \Automaton_n$, whose Parikh vector 
        $\VectorLiteral{m_1, \ldots, m_k}$ satisfies $\phi$.
    \end{tcolorbox}
  }

  The current method of handling this problem consists of computing the
  intersection of the languages (the product of their automata),
  constructing the Presburger representation of their Parikh image, and then
  checking its satisfiability. It is possible to compute an existential
   Presburger formula describing the
  Parikh image of any \emph{context-free} language in linear
  time~\cite{generate-parikh-image}. For our special case of regular languages,
  this result was also stated in~\cite{muscholl-linear}. While theoretically
  elegant, this construction has several disadvantages for us, often making it
  impractical for integration into algorithms.

  Firstly, the constructed Presburger
  formula contains a linear number of existential quantifiers in the size of the
  considered grammar, as well as complex Boolean structure, which is needed to
  express the connectedness of sets of productions considered in the construction.%
  \iffalse{}Eliminating those quantifiers to obtain a quantifier-free representation of the
  Parikh image has exponential complexity~\cite{DBLP:conf/issac/Weispfenning97},
  and is often impossible in reasonable time.\fi{}%
  Just solving the Parikh image
  membership problem is NP-complete, as it corresponds to computing a satisfying
  assignment of the existential Presburger formula, and taxing for solvers as
  well~\cite{ostrich-plus}.

  Secondly, computing the product itself is usually prohibitively expensive
  in both memory and CPU due to exponential blow-up. Applying the approach in~\cite{generate-parikh-image} would in
  this case require the eager computation of the product before its length image,
  and result in an existential Presburger formula of exponential size (in the
  number of automata). In several instances we have observed while solving
  real-world string constraints, the computation of the product of automata
  exhausts the memory of any machine available due to the exponential blow-up in
  size of the product, quickly becoming intractable as the number of automata in
  the product increases.

  The current best published mitigation for this problem relies on the observation
  that any Presburger formula that applies to a product of automata must also hold
  for all of the terms of that product individually. Exploiting this fact, the Parikh
  image of a product of automata is first represented as the conjunction of the image of
  the individual automata of the product, falling back to computing the entire product
  for satisfiable cases~\cite{approximate-parikh}.

@@

Our primary influence for this work is the operation of (finite-domain) 
constraint programming (CP)%
~solvers~\cite{cp}. CP~solvers typically employ a combination of lazy enforcement
of constraints through \emph{propagation},
lazily pruning values from decision variable domains that do not satisfy 
the constraints during solving, and \emph{branching}, splitting proofs
by cases in a way that will encourage maximum propagation.

For our problem, the constraints are automata connectivity and linear
equations (Presburger formulas).
We use propagation to lazily enforce connectivity of automata,
removing transitions as they become unreachable during solving or are
eliminated by constraints in the Presburger formula. We
also branch on the presence or absence of transitions to
encourage propagation of the connectivity constraint. This allows
us to whittle down the automata
(including intermittent products) before moving on with potentially
 explosive product construction. The underlying observation
is that though the problem domain and the domain of our integer variables
is infinite, the number of simple paths
through the automata (and their product) is finite. In summary, the key ideas are:

{
    \centering
    \begin{tcolorbox}[colback=gray!5!white,colframe=gray!75!black,%
        title=Key ideas of \Calculus{},%
        width=0.8\linewidth]

        \begin{itemize}
            \item Enforce automata connectivity constraints lazily
            \item Use \emph{propagation} of constraints on the Parikh image to prune
            transitions from the automata that are incompatible with constraints
            \item Intelligently \emph{branch} on the presence\slash{}absence of key
             transitions to drive propagation of the connectivity constraint
             \item Compute products of automata lazily, after propagation
        \end{itemize}
    \end{tcolorbox}
  }

We implement \Calculus{} as a plug-in theory for the \Princess{} automated
theorem prover.
% Suggestions: l87-89, l102-103: Mention the main evaluation result; such as, x% more than baseline.
%
We additionally wrap the Parikh image solver as a stand-alone tool, \Catra.
\Catra{} also supports a variant of the approximate method of~\cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} distinct Parikh automata intersection problems
generated by \OstrichPlus{} when solving the PyEx string constraint benchmark
suite involving string length constraints~\cite{pyex}.
